A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-06-29 papers/water-SAFT/paper.tex
rename four-rods density files (and edit paper) Tested everything on kipu.

One promising approach for an efficient continuum description of water
is that of classical density-functional theory (DFT), which is an
approach for evaluating the free energy and thermally averaged density
of fluids in an arbitrary external potential~\cite{ebner1976density}.
The foundation of classical DFT is the Mermin
theorem\cite{mermin1965thermal}, which extends the Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} to non-zero temperature,
stating that
\begin{equation}
  A(T) = \underset{n(\rr)}{min}\left\{ F[n(r),T] + \int V_\textit{ext}(\rr) n(\rr)
d\rr\right\}.
\end{equation}
where $A(T)$ is the Helmholtz free energy of a system in the external
potential $V_\textit{ext}$ at temperature $T$, $F[n(r),T]$ is a
universal free-energy functional, independent of external potential,
and $n(\rr)$ is a density of atoms or molecules.  Classical DFT is a
natural framework for creating a more flexible theory of
hydrophobicity that can describe interaction of water with arbitrary
external potentials---including potentials describing strong
interactions with solutes or surfaces.

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, yang1994density, gloor2002saft,gloor2004accurate,
  gloor2007prediction, Jaqaman2004, clark2006developing,
  lischner2010classical, fu2005vapor-liquid-dft,
  blas2001examination}, each of which
captures some of the qualitative behavior of water.  However, each of
these functionals also fail to capture some of water's unique
properties.  For instance, the functional of Lischner \emph{et
  al}\cite{lischner2010classical} treats the surface tension
correctly, but can only be used at room temperature, and thus captures
none of the temperature-dependence of water.  A number of classical
density functionals have recently been produced that are based on
Statistical Associating Fluid Theory (SAFT)\cite{ 
  yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft,gloor2002saft,
  clark2006developing, gloor2007prediction, gloor2004accurate,
  gross2009density, kahl2008modified, blas2001examination}.  These
functionals are based on a perturbative thermodynamic expansion, and
do reproduce the temperature-dependence of water's properties.

@@

SAFT has also been used to construct classical density functionals,
which are commonly used to study the surface tension as a function of
temperature\cite{clark2006developing, gloor2004accurate,
  gloor2007prediction, blas2001examination, gloor2002saft}.  Such
functionals have qualitatively predicted the dependence of surface
tension on temperature, but they also overestimate
the surface tension by about 50\%, and most SAFT-based functionals are
unsuited for studying systems that have density variations on a
molecular length scale due to the use of a local density
approximation\cite{gloor2002saft,clark2006developing, gloor2007prediction,
gloor2004accurate, gross2009density, kahl2008modified, blas2001examination}.

@@

The association functional we use is constructed by using the density
$n_0(\xx)$, which is the density of hard spheres touching a given
point, in the standard SAFT-VR association
energy\cite{gil-villegas-1997-SAFT-VR}.
The association free energy for our four-site model has the form
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx) \zeta(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where the value of $4$ comes from the four association sites per
molecule, and the functional $X$ is the fraction of association sites
\emph{not} hydrogen-bonded, and $\zeta(\xx)$ is a dimensionless
measure of the density inhomogeneity.
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{2V}\cdot\mathbf{n}_{2V}}{n_2^2}.
\end{align}
%
The fraction $X$ is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)
      \Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)
    \Delta(\xx)}
\end{align}
where the functional $\Delta$ which is a measure of hydrogen-bonding
probability, given by
\begin{align}
  \Delta(\xx) &= \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW}
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well attractive
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The correlation function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{HS}_\sigma$, for which we use the
functional of Yu and Wu\cite{yu2002fmt-dft-inhomogeneous-associating}:
\begin{align}
  g_\sigma^{HS} &= \frac{1}{1-n_3}
  +\frac32\frac{\zeta\zeta_2}{(1-n_3)^2}
  + \frac12\frac{\zeta_2^2\zeta}{(1-n_3)^3} \\
  \zeta_2 &= \frac{R}{3}n_2
\end{align}

>>>>

One promising approach for an efficient continuum description of water
is that of classical density-functional theory (DFT), which is an
approach for evaluating the free energy and thermally averaged density
of fluids in an arbitrary external potential~\cite{ebner1976density}.
The foundation of classical DFT is the Mermin
theorem\cite{mermin1965thermal}, which extends the Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} to non-zero temperature,
stating that
\begin{equation}
  A(T) = \underset{n(\rr)}{min}\left\{ F[n(\rr),T] + \int V_\textit{ext}(\rr) n(\rr)
d\rr\right\}.
\end{equation}
where $A(T)$ is the Helmholtz free energy of a system in the external
potential $V_\textit{ext}$ at temperature $T$, $F[n(\rr),T]$ is a
universal free-energy functional, independent of external potential,
and $n(\rr)$ is a density of atoms or molecules.  Classical DFT is a
natural framework for creating a more flexible theory of
hydrophobicity that can describe interaction of water with arbitrary
external potentials---including potentials describing strong
interactions with solutes or surfaces.

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, yang1994density, gloor2002saft, 
  gloor2004accurate, gloor2007prediction, Jaqaman2004, clark2006developing,
  lischner2010classical, fu2005vapor-liquid-dft,kiselev2006new,
  blas2001examination}, each of which
captures some of the qualitative behavior of water.  However, each of
these functionals also fail to capture some of water's unique
properties.  For instance, the functional of Lischner \emph{et
  al}\cite{lischner2010classical} treats the surface tension
correctly, but can only be used at room temperature, and thus captures
none of the temperature-dependence of water.  A number of classical
density functionals have recently been produced that are based on
Statistical Associating Fluid Theory (SAFT)\cite{ 
  yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft,gloor2002saft,muller2001molecular,
  clark2006developing, gloor2007prediction, gloor2004accurate,
  gross2009density, kahl2008modified, blas2001examination}.  These
functionals are based on a perturbative thermodynamic expansion, and
do reproduce the temperature-dependence of water's properties.

@@

SAFT has also been used to construct classical density functionals,
which are commonly used to study the surface tension as a function of
temperature\cite{clark2006developing, gloor2004accurate, kahl2008modified,
  gloor2007prediction, blas2001examination, kiselev2006new,
  gloor2002saft,fu2005vapor-liquid-dft, gross2009density,
  yang1994density}.  Such
functionals have qualitatively predicted the dependence of surface
tension on temperature, but they also overestimate
the surface tension by about 50\%, and most SAFT-based functionals are
unsuited for studying systems that have density variations on a
molecular length scale due to the use of a local density
approximation\cite{gloor2002saft,clark2006developing, gloor2007prediction,
gloor2004accurate, gross2009density, kahl2008modified,
blas2001examination, kiselev2006new}.

@@

The association functional we use is constructed by using the density
$n_0(\xx)$, which is the density of hard spheres touching a given
point, in the standard SAFT-VR association
energy\cite{gil-villegas-1997-SAFT-VR}.
The association free energy for our four-site model has the form
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx) \zeta(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where the value of $4$ comes from the four association sites per
molecule, and the functional $X$ is the fraction of association sites
\emph{not} hydrogen-bonded, and $\zeta(\xx)$ is a dimensionless
measure of the density inhomogeneity.
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{2V}\cdot\mathbf{n}_{2V}}{n_2^2}.
\end{align}
%
The fraction $X$ is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)
      \Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)
    \Delta(\xx)}
\end{align}
where the functional $\Delta$ which is a measure of hydrogen-bonding
probability, given by
\begin{align}
  \Delta(\xx) &= \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW}
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well attractive
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The correlation function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{HS}_\sigma$, for which we use the
functional of Yu and Wu\cite{yu2002fmt-dft-inhomogeneous-associating}:
\begin{align}
  g_\sigma^{HS} &= \frac{1}{1-n_3}
  +\frac{R}{2}\frac{\zeta n_2}{(1-n_3)^2}
  + \frac{R^2}{18}\frac{\zeta n_2^2}{(1-n_3)^3}
\end{align}
